Linting Linearity in Core/System FC

### 2022-11-25 proposal.tex
Cleaning up the work in progress parts

Firstly, we remember the definition of \emph{consuming a value} from Linear
Haskell as
\begin{itemize}
    \item consuming an atomic value is forcing it to Normal Form (NF)
    \item consuming a value that is constructed with more than 1 argument is
        consuming all of its arguments
\end{itemize}
and further note that when an expression is scrutinized by a case expression it
is evaluated to Weak Head Normal Form (WHNF), which in the case of a nullary data
constructor is equal to NF.

Now, with the following example, my interpretation of the current rule says that
the following case expression consumes the scrutinee twice because the case
binder is used. However, we know for sure that in the branch where z is used, z
is equal to False, and using False doesn't violate linearity since we can
unrestrictedly create nullary data constructors. A situation similar to this one
below happens after the CSE transformation.
\begin{code}
    case <complex expression> $\leadsto U$ of z {
        False -> case y of z' { DEFAULT -> z }
        True  -> y
    }
\end{code}

A different example is using the case binder $z$ instead of the arguments we
pattern matched on $x,y$. This currently violates linearity because both $x$ and
$y$ aren't used and because $z$ is used twice. This doesn't typecheck in the
frontend either. However, we know that in this branch, using $z$ is effectively
the same as using $(x,y)$!
\begin{code}
    case <complex expression> $\leadsto U$ of z {
        (x,y) -> z
    }
\end{code}

@@

When we pattern match on a GADT we ...

@@

From the definition of groupBy in GHC.Data.List.Infinite
\begin{code}
groupBy :: (a -> a -> Bool) -> Infinite a -> Infinite (NonEmpty a)
groupBy eq = go
  where
    go (Inf a as) = Inf (a:|bs) (go cs)
      where (bs, cs) = span (eq a) as
\end{code}
we get the following core which violates linearity.
\begin{code}
groupBy = \ (@a) (eq :: a -> a -> Bool) (eta :: Infinite a) ->
  letrec {
    go :: Infinite a -> Infinite (NonEmpty a)
    go = \ (inf :: Infinite a) -> case inf of {
      Inf x xs -> let {
        ds :: ([a], Infinite a)
        ds = let {
            parteq :: a -> Bool
            parteq = eq x
        } in
          letrec {
            goZ :: Infinite a -> ([a], Infinite a)
            goZ = \ (inf' :: Infinite a) -> case wgo inf' of { (# wA, wB #) -> (wA, wB) };

            wgo :: Infinite a -> (# [a], Infinite a #)
            wgo = \ (inf' :: Infinite a) -> case inf' of wildX2 {
              Inf y ys ->
                join {
                    jp :: ([a], Infinite a) \%1 -> (# [a], Infinite a #)
                    jp (wwj :: ([a], Infinite a)) =
                        case wwj of wwj {
                            DEFAULT -> case wwj of { (wA, wB) -> (# wA, wB #) } }
                    } in
                      case parteq y of {
                        False ->
                            let {
                                ww :: ([a], Infinite a)
                                ww = ([] @a, wildX2)
                            } in jump jp ww;
                        True ->
                            let {
                                dy :: ([a], Infinite a)
                                dy = case wgo ys of { (# wA, wB #) -> (wA, wB)
                                }
                            } in
                                let {
                                    wwB :: [a]
                                    wwB = case dy of { (bs, cs) -> bs }
                                } in
                                    let {
                                        wwA :: [a]
                                        wwA = : @a y wwB
                                    } in let {
                                        wwC :: Infinite a
                                        wwC = case dy of { (bs, cs) -> cs }
                                    } in let {
                                        ww :: ([a], Infinite a)
                                        ww = (wwA, wwC)
                                    } in jump jp ww
                                }
              };
          } in
              case wgo xs of { (# wA, wB #) -> (wA, wB) }
      } in
          Inf @(NonEmpty a) (:| @a x (case ds of { (bs, cs) -> bs })) (case ds of { (bs, cs) -> go cs })
    };
  } in go eta
\end{code}
The issue is in the linear function that shows up in the core output (how does
the linearity end up there?). The wwj variable is used once in the case
expression, bound as the case binder, and used in the case body once again. Why
do we do that instead of pattern matching right away? Seems a bit redundant.

Observation: If the branch is DEFAULT, then the case binder binds the case
scrutinee which was just forced?, but hasn't been actually consumed, because we
haven't consumed its components. As long as the same branch doesn't consume the
pattern matching result and the case binder at the same time it should be fine?

In this example, we force wwj with the case binder, but we don't really
consume it (more precise definition of consume...), so we can use the case
binder meaning we're simply using x for the first time. Needs to be
formalised....

@@

Solution with the case binder idea that unifies a different problems:
The case binder is annotated with a usage environment $U_wwj = []$, and its
usage is equal to using the usage environment

Here it would mean that second \textbf{case of} $wwj$ doesn't actually use any
variables and hence it isn't used twice.

@@

@@

\section{Core2Core Validdations}

\subsection{INlining}
\subsection{CSE}

>>>>

Firstly, we remember the definition of \emph{consuming a value} from Linear
Haskell as
\begin{itemize}
    \item Consuming an atomic value is forcing it to Normal Form (NF)
    \item Consuming a value that is constructed with more than 1 argument is
        consuming all of its arguments
\end{itemize}
and further note that when an expression is scrutinized by a case expression it
is evaluated to Weak Head Normal Form (WHNF), which in the case of a nullary data
constructor is equal to NF.

Now, with the following example, my interpretation of the current rule says that
the following case expression consumes the scrutinee twice because the case
binder is used. However, we know for sure that in the branch where z is used, z
is equal to False, and using False doesn't violate linearity since we can
unrestrictedly create nullary data constructors. A situation similar to this one
below happens after the CSE transformation.
\begin{code}
    case <complex expression> of z {
        False -> case y of z' { DEFAULT -> z }
        True  -> y
    }
\end{code}
% $\leadsto U$

A different example is using the case binder $z$ instead of the arguments we
pattern matched on $x,y$. This currently violates linearity because both $x$ and
$y$ aren't used and because $z$ is used twice. This doesn't typecheck in the
frontend either. However, we know that in this branch, using $z$ is effectively
the same as using $(x,y)$!
\begin{code}
    case <complex expression> of z {
        (x,y) -> z
    }
\end{code}
% $\leadsto U$

@@

% When we pattern match on a GADT we ...

@@

From the definition of $\mathit{groupBy}$ in $\mathit{GHC.Data.List.Infinite}$:

\begin{code}
groupBy :: (a -> a -> Bool) -> Infinite a -> Infinite (NonEmpty a)
groupBy eq = go
  where
    go (Inf a as) = Inf (a:|bs) (go cs)
      where (bs, cs) = span (eq a) as
\end{code}

We get a somewhat large resulting core expression, in the middle of which the
following expression with a linear type -- which currently syntatically violates
linearity.

@@

Using the case binder usage solution, the case binder is annotated with a usage
environment for the only branch ($U_wwj = \emptyset$) and using $wwj$ is equal
to using the $\emptyset$. Here it would mean that second \textbf{case of} $wwj$
doesn't actually use any variables and hence the first $wwj$ isn't used twice.

% \begin{code}
% groupBy = \ (@a) (eq :: a -> a -> Bool) (eta :: Infinite a) ->
%   letrec {
%     go :: Infinite a -> Infinite (NonEmpty a)
%     go = \ (inf :: Infinite a) -> case inf of {
%       Inf x xs -> let {
%         ds :: ([a], Infinite a)
%         ds = let {
%             parteq :: a -> Bool
%             parteq = eq x
%         } in
%           letrec {
%             goZ :: Infinite a -> ([a], Infinite a)
%             goZ = \ (inf' :: Infinite a) -> case wgo inf' of { (# wA, wB #) -> (wA, wB) };

%             wgo :: Infinite a -> (# [a], Infinite a #)
%             wgo = \ (inf' :: Infinite a) -> case inf' of wildX2 {
%               Inf y ys ->
%                 join {
%                     jp :: ([a], Infinite a) \%1 -> (# [a], Infinite a #)
%                     jp (wwj :: ([a], Infinite a)) =
%                         case wwj of wwj {
%                             DEFAULT -> case wwj of { (wA, wB) -> (# wA, wB #) } }
%                     } in
%                       case parteq y of {
%                         False ->
%                             let {
%                                 ww :: ([a], Infinite a)
%                                 ww = ([] @a, wildX2)
%                             } in jump jp ww;
%                         True ->
%                             let {
%                                 dy :: ([a], Infinite a)
%                                 dy = case wgo ys of { (# wA, wB #) -> (wA, wB)
%                                 }
%                             } in
%                                 let {
%                                     wwB :: [a]
%                                     wwB = case dy of { (bs, cs) -> bs }
%                                 } in
%                                     let {
%                                         wwA :: [a]
%                                         wwA = : @a y wwB
%                                     } in let {
%                                         wwC :: Infinite a
%                                         wwC = case dy of { (bs, cs) -> cs }
%                                     } in let {
%                                         ww :: ([a], Infinite a)
%                                         ww = (wwA, wwC)
%                                     } in jump jp ww
%                                 }
%               };
%           } in
%               case wgo xs of { (# wA, wB #) -> (wA, wB) }
%       } in
%           Inf @(NonEmpty a) (:| @a x (case ds of { (bs, cs) -> bs })) (case ds of { (bs, cs) -> go cs })
%     };
%   } in go eta
% \end{code}
% The issue is in the linear function that shows up in the core output (how does
% the linearity end up there?). The wwj variable is used once in the case
% expression, bound as the case binder, and used in the case body once again. Why
% do we do that instead of pattern matching right away? Seems a bit redundant.

% Observation: If the branch is DEFAULT, then the case binder binds the case
% scrutinee which was just forced?, but hasn't been actually consumed, because we
% haven't consumed its components. As long as the same branch doesn't consume the
% pattern matching result and the case binder at the same time it should be fine?

% In this example, we force wwj with the case binder, but we don't really
% consume it (more precise definition of consume...), so we can use the case
% binder meaning we're simply using x for the first time. Needs to be
% formalised....

@@

\section{Work in progress}

% \section{Core to Core Validation}
\subsection{Inlining}

\subsection{CSE}

@@
